---
title: "Analysis of NOAA Storm Database: Economic and human costs of severe weather"
output: 
  html_document:
    keep_md: true
---

## Synopsis 

The goal of this data analysis was to address the following questions:

1.	Across the United States, which types of weather events are most harmful with respect to population health?
1.	Across the United States, which types of weather events have the greatest economic consequences?

This document describes the steps taken in the data processing of the NOAA Strom Database, as well as the analysis of the results.  In short, two types of severe weather events stick out, Tornado because of their high rate of fatalities, and flooding for the economic costs of crop and property damage.


## Libraries

```{r, warning=FALSE}
#load libraries
library(ggplot2)
```

2) Download and unpack the dataset

```{r cache=TRUE}
# download file
if (!file.exists("StormData.csv.bz2")) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
                ,"StormData.csv.bz2")
}

storm_data <- read.csv(bzfile('StormData.csv.bz2'), header=TRUE, stringsAsFactors = FALSE)
```

3) Group the fatalities column by event type and sum.  Then restrict the dataset to the top ten results 

```{r}
reduced_fatal = aggregate(storm_data$FATALITIES,by=list(storm_data$EVTYPE),FUN=sum, na.rm=TRUE)
result_fatal <- head(reduced_fatal[order(reduced_fatal$x, decreasing=TRUE),], 10)
```

4) Group the injuries column by event type and sum.  Then restrict the dataset to the top ten results

```{r}
reduced_injury = aggregate(storm_data$INJURIES,by=list(storm_data$EVTYPE),FUN=sum, na.rm=TRUE)
result_injury <- head(reduced_injury[order(reduced_injury$x, decreasing=TRUE),], 10)
```

5) Organize the NOAA Strom data set to calculate total ecenomic impact, set variables used for processing

```{r}
exp <- c("B", "b", "M", "m", "K", "k", "H", "h")
exp_v <- c( H=10^2, h=10^2, M=10^6, m=10^6, K=10^3, k=10^3, B=10^9, b= 10^9)
titles <- c("EventType", "Damage", "Multipler")
```

6) Subset the property damage by the property exponent, and rename columns

```{r}
prop_set <- subset(storm_data, storm_data$PROPDMGEXP %in% exp)
prop_tmp <- prop_set[, c("EVTYPE", "PROPDMG", "PROPDMGEXP")]
colnames(prop_tmp) <- titles
```

7) Subset the crop damage by the crop exponent, and rename columns

```{r}
crop_set <- subset(storm_data, storm_data$CROPDMGEXP %in% exp)
crop_tmp <- crop_set[, c("EVTYPE", "CROPDMG", "CROPDMGEXP")]
colnames(crop_tmp) <- titles
```

8) Bind the Crop and Property datasets and calculate the dollar amount (total_dollars) 

```{r}
tmp_result <- rbind(crop_tmp, prop_tmp)
tmp_result$total_dollars <- tmp_result$Damage * exp_v[tmp_result$Multipler]
```

9) Group the damages by event type and sum.  Then restrict the dataset to the top ten results 

```{r}
reduced_damage = aggregate(tmp_result$total_dollars ,by=list(tmp_result$EventType),FUN=sum)
result_damage <- head(reduced_damage[order(reduced_damage$x, decreasing=TRUE),], 10)
```

## Results

### Qusetion One:	Across the United States, which types of weather events are most harmful with respect to population health?

The two charts that follow organise the total fatalities and injuries by weather event. In both cases Tornado's stand out, 
as the most hazzardous event.

```{r}
ggplot(data=result_fatal, aes(x=reorder(result_fatal$Group.1, result_fatal$x), y=result_fatal$x)) + 
  geom_bar(stat="identity") + xlab("Weather Event") + ylab("Total Fatalities") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
```

```{r}
ggplot(data=result_injury, aes(x=reorder(result_injury$Group.1, result_injury$x), y=result_injury$x)) + 
  geom_bar(stat="identity") + xlab("Weather Event") + ylab("Total Injuries") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
```

### Qusetion Two:	Across the United States, which types of weather events have the greatest economic consequences?

The chart that follows combines the total cost of crop and property damage incurred by weather events.  Flooding has been responsible for over 150 billion dollars in damage, suggesting it has the largest economic impact of all the weather events.

```{r}
ggplot(data=result_damage, aes(x=reorder(result_damage$Group.1, result_damage$x), y=result_damage$x)) + 
  geom_bar(stat="identity") + xlab("Weather Event") + ylab("Total Damages (US dollars)") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
```

### October, 2015
